---
title: "A3: Incarceration"
author: "Sára Mustre-del Río"
output: html_document
---

### Introduction

It is a widespread fact that incarceration rates have historically been disproportionately higher for Black people relative to other racial groups in the U.S. This trend notably rose during the Reagan administration, which was known for being “tough on crime” and disproportionately targeting lower-income Black and brown communities. However, President Barack Obama’s election in 2009 brought hope for progress against racial inequities, given that he was the first Black president. Because of this, I wanted to analyze incarceration trends across the U.S. during his administration, which ranged from 2009 to 2018. 

### Summary Information

In the first section, I analyze total prison admissions for both Black and white populations, and I also highlight differences in gender (1a). I also analyze the average prison admissions for the abovementioned populations, but I focus on the different geographical regions, such as the Midwest, West, Northeast, and  the South (1b).

In the second section, I focus on total admissions per year, dividing them between Black (2a) and white (2b) populations. California had some of the highest admission counts for both Black and white populations, and I assume that is because of the population density. Other counties that had high counts of incarceration for both populations had big cities in them and were located in the South and West areas. This might have to do with more diverse populations and unequal wealth distributions within large urban settings.

In the third section, I calculate the abovementioned values‘ averages. While California leads in both, states with higher Black populations—such as  Maryland—remain towards the top (3a). States with higher white populations, such as Rhode Island, which is almost 90% white, have the highest prison admissions (3b).

In the fourth section, I analyze Black and white admission counts by region. In total counts, States that have high Black populations generally top the charts. However, I was surprised with southern statistics, as states with higher Black populations, such as Mississippi, Louisiana and Georgia, do not have such high counts (4a). Averages, however, reflect many of the observations I made in the graphs (4b).

### The Dataset

The Vera Project collected and condensed this data from many different sources. This is a good way to ensure that the most data possible is collected and the least amount possible of gaps are left unfilled. However, it is important to know how data is collected, as well as how different categorical variables might be defined, such as race and gender. For example, does the gender category take into account how different states define transgender people?

I think that while this data is relevant, I would have better worked with proportions, as this data does not contain the populations of each race in a given county or state. For example, it makes sense that California has some of the highest counts of prison admissions in both racial groups, given that it is a very densely populated state, even if it has less than 6% Black population.

Another limitation I found was how gender was displayed in the data. I really wanted to find a way to display gender differences in admission rates in my graphs, but I would have found this more practical if the data was organized differently. I think having a separate column that classifies admission rates by gender would have been more helpful. That way, I could have turned gender into a variable in my data.

Other information that could have been useful would be the most common reasons for incarceration for both groups. In different Law, Societies and Justice classes—which is my major—I have learned a lot about how Black and brown people are more likely to receive harsher punishment for minor misdemeanors or crimes committed by necessity, such as stealing. 


### Trends Over Time Chart

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)

incarceration_df<-read.csv("https://raw.githubusercontent.com/vera-institute/incarceration-trends/master/incarceration_trends.csv")

avg_black_adm<-incarceration_df %>% 
  group_by(year, state, region) %>% 
  filter(year > 2008) %>% 
  select(year, region, state, county_name, black_prison_adm, black_male_prison_adm, black_female_prison_adm) %>% 
  summarize(avg_black_adm = mean(black_prison_adm, na.rm=TRUE),.groups = 'drop')

custom_colors<- c("#8DCE67", "#6DAB6C", "#73B882", "#64AC88", "#71C9C3")

ggplot(avg_black_adm) +
  geom_col(mapping = aes(x = year, y = avg_black_adm, fill = region), position = "dodge")+
  labs(title = 'Average Black incarceration during the Obama Administration',
       subtitle = 'by U.S. geographic region',
       x = 'Year',
       y = 'Average Black prisoner admissions',
       fill = 'Region')+
  scale_fill_manual(values = custom_colors)
```

This column shows average admissions of Black prisoners during the Obama administration. The Midwest’s admission counts dramatically decreased, which might tell us about how the election of a Black president might have influenced states’ policing practices.

### Variable Comparison Chart

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)

incarceration_df<-read.csv("https://raw.githubusercontent.com/vera-institute/incarceration-trends/master/incarceration_trends.csv")

prison_adm_total<-incarceration_df %>% 
  group_by(year) %>% 
  filter(year > 2008) %>%
  select(year, region, state, black_prison_adm, black_male_prison_adm,black_female_prison_adm, 
         white_prison_adm,white_female_prison_adm,white_male_prison_adm) %>% 
  drop_na()

ggplot(data = prison_adm_total) +
  geom_point(
    mapping = aes(x = white_prison_adm, y = black_prison_adm, color = region),
    alpha = .6
  ) + 
  facet_wrap(~year)+
  scale_color_brewer(palette = "PiYG")+
  labs(title = 'Total Black vs white incarceration during the Obama administration',
       subtitle = 'by U.S. geographic region',
       x = 'Total admissions of white prisoners',
       y = 'Total admissions of Black prisoners',
       fill = 'region')
```

Here, I compare the total admissions of Black vs white prisoners during the Obama administration by U.S. geographic region. These have been somewhat consistent throughout the years except for higher prison admission counts of white people in the West, and higher admission counts for Black people in the Northeast and South.

### Map

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library('tidyverse')
library(ggplot2)
library(dplyr)
library(tidyr)

incarceration_df<-read.csv("https://raw.githubusercontent.com/vera-institute/incarceration-trends/master/incarceration_trends.csv")

state_shape <- map_data('state')

state_abbrevs <- data.frame(state.abb, state.name)

total_black_adm<-incarceration_df %>% 
  group_by(year) %>% 
  filter(year > 2008) %>%
  select(year,region, state, black_prison_adm) %>% 
  drop_na()

sum_total_black_adm <- total_black_adm %>% 
  group_by(state) %>% 
  summarize(state_total=sum(black_prison_adm))

sum_total_black_adm <- left_join(sum_total_black_adm, state_abbrevs, by=c('state' = 'state.abb'))

sum_total_black_adm <- sum_total_black_adm %>% 
  mutate(state.name=tolower(state.name))

state_shape_2<- left_join(sum_total_black_adm, state_shape, by=c('state.name'='region')) 

ggplot(state_shape_2)+
  geom_polygon(mapping=aes(x=long, y=lat, group=group, fill=state_total))+
  coord_map()+
  labs(title = 'Average admissions of Black prisoners during the Obama administration', fill = 'state_total')+
  scale_fill_continuous(low='grey', high='black',labels=scales::label_number_si())

```

This map shows which states had the highest average admissions for Black prisoners during the Obama administration. Texas had the highest average, followed by California.