---
title: "A3: Incarceration"
output: html_document
---

### Introduction

It is a widespread fact that incarceration rates have historically been disproportionately higher for Black people relative to other racial groups in the U.S. This trend notably rose during the Reagan administration, which was known for being “tough on crime” and disproportionately targetting lower-income Black and brown communities. However, President Barack Obama’s election in 2009 brought hope for progress against racial inequities, given that he was the first Black president. Because of this, I wanted to analyze incarceration trends across the U.S. during his administration, which ranged from 2009 to 2018. 

### Summary Information

```{r summary info, eval=FALSE, message=FALSE}
source("Summary.R")
```

### The Dataset

The Vera Project collected and condensed this data from many different sources. This is a good way to ensure that the most data possible is collected and the least amount possible of gaps are left unfilled. However, it is important to know how data is collected, as well as how different categorical variables might be defined, such as race and gender. For example, does the gender category take into account how different states define transgender people?

I think that while this data is relevant, I would have better worked with proportions, as this data does not contain the populations of each race in a given county or state. For example, it makes sense that California has some of the highest counts of prison admissions in both racial groups, given that it is a very densely populated state, even if it has less than 6% Black population.

Another limitation I found was how gender was displayed in the data. I really wanted to find a way to display gender differences in admission rates in my graphs, but I would have found this more practical if the data was organized differently. I think having a separate column that classifies admission rates by gender would have been more helpful. That way, I could have turned gender into a variable in my data.

Other information that could have been useful would be the most common reasons for incarceration for both groups. In different Law, Societies and Justice classes—which is my major—I have learned a lot about how Black and brown people are more likely to receive harsher punishment for minor misdemeanors or crimes committed by necessity, such as stealing. 


### Trends Over Time Chart

``````{r, echo = FALSE, code = readLines("chart1_example.R")}
```

### Variable Comparison Chart

```{r, echo = FALSE, code = readLines("chart2_example.R")}
```

### Map

```{r, echo = FALSE, code = readLines("map_example.R")}
```
